An introduction to Docker for reproducible research

### 2014-12-04 docker-paper/manuscript.tex
archiving submitted copies

---
title: "An introduction to Docker for reproducible research"
author: 
  - name: Carl Boettiger
    affiliation:
      l1: Center for Stock Assessment Research, 
      l2: 110 Shaffer Rd, Santa Cruz, CA 95050, USA 
    email: cboettig(at)gmail.com
    footnote: Corresponding author
abstract: |
  As computational work becomes more and more integral to many aspects of
  scientific research, computational reproducibility has become an issue
  of increasing importance to computer systems researchers and domain
  scientists alike. Though computational reproducibility seems more
  straight forward than replicating physical experiments, the complex
  and rapidly changing nature of computer environments makes being able
  to reproduce and extend such work a serious challenge.  In this paper,
  I explore common reasons that code developed for one research project
  cannot be successfully executed or extended by subsequent researchers.
  I review current approaches to these issues, including virtual machines
  and workflow systems, and their limitations. I then examine how the
  popular emerging technology Docker combines several areas from systems
  research - such as operating system virtualization, cross-platform
  portability, modular re-usable elements, versioning, and a 'DevOps'
  philosophy, to address these challenges.  I illustrate this with several
  examples of Docker use with a focus on the R statistical environment.

Keywords:
  - Docker
  - virtualization
  - reproducibility

bibliography: components/references.bib
csl: components/acm-sig-proceedings.csl
citecolor: black
linkcolor: black
urlcolor: black

output: 
  pdf_document:
    template: components/sigproc-sp.tex
    keep_tex: true
    fig_caption: true
---

<!-- Excerpt from the CALL FOR PAPERS:
There is increasing interest in promoting repeatable and reproducible research
in computer science.  A community gains confidence in scientific claims when
the experiments supporting those claims are published, examined, and repeated.
Repeating research also helps a community expand on previous results and
discover new directions for investigation.  Repetition requires that the
artifacts used in an experiment be made available to, and made usable by,
investigators.  In other words, artifacts must be shared.  Possible artifacts
include not only binaries, source code, and datasets but also computing and
networking infrastructure such as testbeds and clouds.

Systems researchers face special challenges in the areas of repeatability and
artifact sharing.  Some arise from the nature of systems software, which is
typically complex and rife with internal and external dependencies.  Other
issues arise from execution environments: for example, to reproduce performance
and scalability results, one may need access to actual and specific hardware.
Still more systems-specific challenges arise from the difficulty of measuring
properties of interest, such as time and energy consumption.

Suggested paper topics include, but are not limited to:

  * novel ways to manage systems research artifacts, dependencies, workflows,
    and data collection
  * testbeds for repeatable research and/or publishing executable artifacts
  * "rich publications" that connect papers to software and data
  * repeatability in distributed and nondeterministic systems
  * mitigating obstacles to artifact sharing, repeatability, and reproduction
  * repeatability challenges specific to systems
  * measurement bias in systems experiments
  * experiences and outcomes from systems repetition and reproduction studies
  * improving scientific practice in systems research
  * incentives for producing and sharing high-quality research artifacts
-->

# Introduction

Reproducible research has received an increasing level of attention
throughout the scientific community [_e.g._ @Peng_2011; @Nature_2012]
and the public at large [_e.g._ @The_Economist_2013]. All steps of the
scientific process, from data collection and processing, to analyses,
visualizations and conclusions depend ever more on computation and
algorithms, _computational reproducibility_ has received particular
attention [@Merali_2010].  Though in principle this algorithmic dependence
should make such research easier to reproduce -- computer codes being
both more portable and potentially more precise to exchange and run
than experimental methods -- in practice this has led to an ever larger
and more complex black box that stands between what was actually done
and what is described in the literature.  Crucial scientific processes
such as replicating the results, extending the approach or testing the
conclusions in other contexts, or even merely installing the software
used by the original researchers can become immensely time-consuming if
not impossible.

### Systems research & reproducibility

Systems research has long concerned itself with the issues of
computational reproducibility and the technologies that can facilitate
those objectives [_e.g._ @Eide_2010, @Harji_2013]. Docker is a new but already
very popular open source tool that combines many of these approaches
in a user friendly implementation, including: (1) performing Linux
container (LXC) based operating system (OS) level virtualization, (2)
portable deployment of containers across platforms, (3) component reuse,
(4) sharing, (5) archiving, and (6) versioning of container images. While
Docker's market success has largely focused on the needs of businesses
in deploying web applications and the potential for a lightweight
alternative to full virtualization, these features have potentially
important implications for systems research in the area of scientific
reproducibility.

In this paper, I seek to address two audiences. First, that of the domain
scientist, those conducting research in ecology, bioinformatics, economics,
psychology and so many other disciplines in which computation plays an 
ever-increasing role. I seek to help this audience become more aware of the 
concerns and challenges in making these computations more reproducible, extensible, 
and portable to other researchers, and highlight Docker as platform to address 
these challenges. The second audience is that of the computer systems researchers:
readers familiar with both the challenges to reproducibility (versioning,
software dependencies, etc.) and the technical components underlying Docker (virtualization,
LXC containers, jails, hashes, etc.), but who may be unfamiliar with Docker software. 
I hope the domain scientist will be motivated to try Docker as a tool to address
reproducibility. Meanwhile, I expect the systems researcher to see Docker technology as
an area ripe for further systems research (e.g. to what extent can differences between 
Linux kernels frustrate reproducibility in a Docker image?) rather than as a tool
for their own study (because Docker applies only to software at the level above
the Linux kernel, it will be of little use in making studies of differences in 
kernel behavior, hardware performance, execution times, and so forth reproducible.)

### A cultural problem

It is worth observing from the outset that the primary barrier to
computational reproducibility in many domain sciences has nothing
to do with the technological approaches discussed here, but stems rather from a
reluctance to publish the code used in generating the results in the
first place [@Barnes_2010]. Despite extensive evidence to the contrary
[_e.g._ @Ince_2012], many researchers and journals continue to assume that
summary descriptions or pseudo-code provide a sufficient description of
computational methods used in data gathering, processing, simulation,
visualization, or analysis.  Until such code is available in the first
place, one cannot even begin to encounter the problems that the approaches
discussed here set out to solve. As a result, few domain researchers
may be fully aware of the challenges involved in effectively re-using
published code.

A lack of requirements or incentives no doubt plays a crucial role in
discouraging sharing [@Barnes_2010; @Stodden_2013]. Nevertheless,
it is easy to underestimate the significant barriers raised by a lack
of familiar, intuitive, and widely adopted tools for addressing the
challenges of computational reproducibility. Surveys and case studies find
that a lack of time, more than innate opposition to sharing, discourages
researchers from providing code [@Stodden_2010; @FitzJohn_2014].

## Four technical challenges

@@

Studies focusing on code that has been made available with scientific
publications regularly find the same common issues that pose substantial
barriers to reproducing the original results or building on that code
[_e.g._ @Lapp_2014; @Garijo_2013; @Gilbert_2012; @Collberg_2014;
@Krishnamurthi_2014], which I attempt to summarize as follows.

### 1. **"Dependency Hell"**

A recent study by researchers at the University of Arizona found that
less than 50% of software could even be successfully built or installed
[@Collberg_2014]. Though sufficiently knowledgeable users may be able
to overcome the issues in at least some of these cases [^6], similar
results are seen in an ongoing effort by other researchers to replicate
that study [@Krishnamurthi_2014], and has also been observed in other
indepedent studies [@Lapp_2014; @Garijo_2013].  Installing or building
software necessary to run the code in question assumes the ability to
recreate the computational environment of the original researchers.

[^6]: See discussion at https://gist.github.com/pbailis/9629050 and https://gist.github.com/samth/9641364

Differences in numerical evaluation, such as arise in floating point
arithmetic or even ambiguities in standardized programming
languages ("order-of-evaluation" problems) can be responsible for
differing results between or even within the same computational platform
[@Ince_2012]. Such issues make it difficult to restrict the true dependencies
of the code to higher level environments such as that of a given scripting
language, independent of the underlying OS or even hardware itself.

### 2.  **Imprecise documentation**

Documentation on how to install and run code associated with published
research is another frequent barrier to replication. A study by Lapp [@Lapp_2014] found
this impairs a researcher's ability to install and build the software
necessary, as even small holes in the documentation were found to be
major barriers, particularly for "novices" [Sensu @Garijo_2013] --
where novices may be experts in nearby languages but unfamiliar with
the package managers and other tools of the language involved. This
same problem is discussed in [@Clark_2014].  Imprecise documentation
goes well beyond issues of the software environment itself: incomplete
documentation of parameters involved meant as few as 30% of analyses
($n=34$) using the popular software STRUCTURE could be reproduced in the
study of [@Gilbert_2012].

### 3.  **Code rot**

Software dependencies are not static elements, but receive regular
updates that may fix bugs, add new features or deprecate old features
(or even entire dependencies themselves).  Any of these changes can
potentially change the results generated by the code.  As some of these
changes may indeed resolve valid bugs or earlier problems with underlying
code, it will often be insufficient to demonstrate that results can be
reproduced when using the original versions, a problem sometimes known
as "code rot."  Researchers will want to know if the results are robust
to the changes. The case studies in [@Lapp_2014] provide examples of
these problems.

### 4. **Barriers to adoption and reuse in existing solutions**

Technological solutions such as workflow software, virtual machines,
continuous integration services, and best practices from software
development would address many of the issues frequently frustrating
reproducibility. However, researchers face significant barriers to entry
in learning these tools and approaches which are not part of their typical
curriculum, or lack incentives commensurate with the effort required
[@Joppa_2013; @FitzJohn_2014].

<!-- Obviously reproducible research barriers go much deeper than this,
code that is not modular, not documented, not functionalized or otherwise
easy to read and follow are all obstacles to resuse.-->

<!-- A wide range of proposals, I focus on those at the operating system level -->
Though a wide variety of approaches exists to work around these challenges,
few operate on a low enough level to provide a general solution. Clark _et al._ [@Clark_2014]
provide an excellent description of this situation:

> In scientific computing the environment was commonly managed via
Makefiles & Unix-y hacks, or alternatively with monolithic software
like Matlab. More recently, centralized package management has provided
curated tools that work well together. But as more and more essential
functionality is built out across a variety of systems and languages,
the value -- and also the difficulty -- of coordinating multiple tools
continues to increase. Whether we are producing research results or web
services, it is becoming increasingly essential to set up new languages,
libraries, databases, and more.

@@

# Current approaches

<!-- FIXME Doesn't say much about the approaches. Not clear what a workflow is... -->

Two dominant paradigms have emerged to address these issues so far:
workflow software [_e.g._ @Altintas_2004; @Hull_2006]  and
virtual machines [_e.g._ @Dudley_2010; @Howe_2012]. 
Workflow software provides very elegant technical solutions to the
challenges of communication between diverse software tools, capturing
provenance in graphically driven interfaces, and handling issues from 
versioning dependencies to data access.  Workflow solutions are often 
built by well-funded collaborations between domain scientists and computer
scientists, and can be very successful in the communities within which 
they receive substantial adoption. Nonetheless, most workflow systems 
struggle with relatively low total adoption overall [@Gil_2007; @Dudley_2010].

Dudley & Butte [@Dudley_2010] give several reasons that such comprehensive workflow
systems have not been more successful:

> (i) efforts are not rewarded by the current academic research and
funding environment; (ii) commercial software vendors tend to protect
their markets through proprietary formats and interfaces; (iii)
investigators naturally tend to want to 'own' and control their
research tools; (iv) even the most generalized software will not be
able to meet the needs of every researcher in a field; and finally (v)
the need to derive and publish results as quickly as possible precludes
the often slower standards-based development path.

In short, workflow software expects a new approach to computational research. 
In contrast, virtual machines (VMs) offer a more direct approach. Since
the computer Operating System (OS) already provides the software layer
responsible for coordinating all the different elements running on the computer,
the VM approach captures the OS and everything running on it whole-cloth.
To make this practical, Dudley & Butte [@Dudley_2010] and Howe [@Howe_2012] both
propose using virtual machine images that will run on the cloud, such as Amazon's
EC2 system, which is already based upon this kind of virtualization.

Critics of the use of VMs to support reproducibility highlight that the 
approach is too much of a black box and thus ill suited for reproducibility [@Watson_2012].
While the approach sidesteps the need to either install or even document the
dependencies, this also makes it more difficult for other researchers to
understand, evaluate, or alter those dependencies.  Moreover, other research
cannot easily build on the virtual machine in a consistent and scalable way. 
If each study provided its own virtual machine, any pipeline combining the 
tools of multiple studies would quickly become impractical or impossible 
to implement.

<!-- Scripted provisioning: a solution to the weaknesses of virtual machines -->
## A "DevOps" approach

The problems highlighted here are not unique to _academic_ software,
but impact software development in general. While the academic research
literature has frequently focused on the development of workflow software
dedicated to particular domains, or otherwise to the use of virtual
machines, the software development community has recently emphasized a
philosophy (rather than a particular tool), known as _Development_ and
Systems _Operation_, or more frequently just "DevOps." The approach is
characterized by scripting, rather than documenting, a description of the
necessary dependencies for software to run, usually from the Operating
System (OS) on up. Clark _et al._ [@Clark_2014] describe the DevOps approach along
with both its relevance to reproducible research and examples of its
use in the academic research context. They identify the difficulties
I have discussed so far in terms of effective documentation:

> Documentation for complex software environments is stuck between two
opposing demands. To make things easier on novice users, documentation
must explain details relevant to factors like different operating
systems. Alternatively, to save time writing and updating documentation,
developers like to abstract over such details.

The authors contrast this to the DevOps approach, where dependency 
documentation is _scripted_:

> A DevOps approach to "documenting" an application might consist of
providing brief descriptions of various install paths, along with scripts
or "recipes" that automate setup.

This elegantly addresses both the demand for simplicity of use (one
executes a script instead of manually managing the environmental setup)
and comprehensiveness of implementation. Clark _et al._ [@Clark_2014] are careful to note
that this is not so much a technological shift as a philosophical one:

> The primary shift that's required is not one of new tooling, as most
developers already have the basic tooling they need. Rather, the needed
shift is one of philosophy.

Nevertheless, a growing suite of tools designed explicitly for this
purpose have rapidly replaced the use of general purpose tools (such
as Makefiles, bash scripts) to become synonymous with the DevOps
philosophy.  Clark _et al._ [@Clark_2014] reviews many of these DevOps tools, their
different roles, and their application in reproducible research.

I focus the remainder of this paper on one of the most recent and
rapidly growing among these, called Docker, and the role it can play
in reproducible research. Docker offers several promising features
for reproducibility that go beyond the tools highlighted in [@Clark_2014].
Nevertheless, my goal in focusing on this technology is not to
promote a particular solution, but to anchor the discussion of technical
solutions to reproducibility challenges in concrete examples.

<!-- Link on how to install Docker -->
<!-- Examples here are not meant as a tutorial to Docker, but 
rather to illustrate concepts a reader should be able to follow whether 
or not they have any previous knowledge of Docker and its commands -->

# Docker

<!-- Docker is an open source (Apache Version 2) software platform written in
the Go programming language.-->

>>>>

% THIS IS SIGPROC-SP.TEX - VERSION 3.1
% WORKS WITH V3.2SP OF ACM_PROC_ARTICLE-SP.CLS
% APRIL 2009
%
% It is an example file showing how to use the 'acm_proc_article-sp.cls' V3.2SP
% LaTeX2e document class file for Conference Proceedings submissions.
% ----------------------------------------------------------------------------------------------------------------
% This .tex file (and associated .cls V3.2SP) *DOES NOT* produce:
%       1) The Permission Statement
%       2) The Conference (location) Info information
%       3) The Copyright Line with ACM data
%       4) Page numbering
% ---------------------------------------------------------------------------------------------------------------
% It is an example which *does* use the .bib file (from which the .bbl file
% is produced).
% REMEMBER HOWEVER: After having produced the .bbl file,
% and prior to final submission,
% you need to 'insert'  your .bbl file into your source .tex file so as to provide
% ONE 'self-contained' source file.
%
% Questions regarding SIGS should be sent to
% Adrienne Griscti ---> griscti@acm.org
%
% Questions/suggestions regarding the guidelines, .tex and .cls files, etc. to
% Gerald Murray ---> murray@hq.acm.org
%
% For tracking purposes - this is V3.1SP - APRIL 2009

\documentclass[9pt]{components/acm_proc_article-sp}
%\documentclass{acm_proc_article-sp}
\usepackage[hyphens]{url}
\usepackage{graphicx}
\usepackage{booktabs} % book-quality tables

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}

\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi

  % use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{{#1}}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{{#1}}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{{#1}}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{{#1}}}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{{#1}}}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{{#1}}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{{#1}}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{{#1}}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{{#1}}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{{#1}}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{{#1}}}
\newcommand{\RegionMarkerTok}[1]{{#1}}
\newcommand{\ErrorTok}[1]{\textbf{{#1}}}
\newcommand{\NormalTok}[1]{{#1}}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={An introduction to Docker for reproducible research},
            colorlinks=true,
            urlcolor=black,
            linkcolor=black,
            citecolor=black,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}
% Pandoc toggle for numbering sections (defaults to be off)
\setcounter{secnumdepth}{0}
% Pandoc header

 %% Omit the copyright space.
%  \makeatletter
%  \def\@copyrightspace{}
%  \makeatother

  %% Fill in the title and author list for your paper.
%  \title{An introduction to Docker for reproducible research}
%  \author{ \thanks{Copyright is held by the author(s)}}
%  \renewcommand\footnotemark{}

\toappear{Copyright is held by the author(s)}
\begin{document}

\title{An introduction to Docker for reproducible research}
\subtitle{}
%
% You need the command \numberofauthors to handle the 'placement
% and alignment' of the authors beneath the title.
%
% For aesthetic reasons, we recommend 'three authors at a time'
% i.e. three 'name/affiliation blocks' be placed beneath the title.
%
% NOTE: You are NOT restricted in how many 'rows' of
% "name/affiliations" may appear. We just ask that you restrict
% the number of 'columns' to three.
%
% Because of the available 'opening page real-estate'
% we ask you to refrain from putting more than six authors
% (two rows with three columns) beneath the article title.
% More than six makes the first-page appear very cluttered indeed.
%
% Use the \alignauthor commands to handle the names
% and affiliations for an 'aesthetic maximum' of six authors.
% Add names, affiliations, addresses for
% the seventh etc. author(s) as the argument for the
% \additionalauthors command.
% These 'additional authors' will be output/set for you
% without further effort on your part as the last section in
% the body of your article BEFORE References or any Appendices.

\numberofauthors{1} %  in this sample file, there are a *total*
% of EIGHT authors. SIX appear on the 'first-page' (for formatting
% reasons) and the remaining two appear in the \additionalauthors section.
%
\author{
% You can go ahead and credit any number of authors here,
% e.g. one 'row of three' or two rows (consisting of one row of three
% and a second row of one, two or three).
%
% The command \alignauthor (no curly braces needed) should
% precede each author name, affiliation/snail-mail address and
% e-mail address. Additionally, tag each line of
% affiliation/address with \affaddr, and tag the
% e-mail address with \email.
%
% 1st. author
  \alignauthor
  Carl Boettiger %  \titlenote{Corresponding author}
  \\
  \affaddr{\small Center for Stock Assessment Research,}\\
  \affaddr{\small 110 Shaffer Rd, Santa Cruz, CA 95050, USA}\\
  \email{\small cboettig(at)gmail.com}
}
\date{\today}

\maketitle
\begin{abstract}
As computational work becomes more and more integral to many aspects of
scientific research, computational reproducibility has become an issue
of increasing importance to computer systems researchers and domain
scientists alike. Though computational reproducibility seems more
straight forward than replicating physical experiments, the complex and
rapidly changing nature of computer environments makes being able to
reproduce and extend such work a serious challenge. In this paper, I
explore common reasons that code developed for one research project
cannot be successfully executed or extended by subsequent researchers. I
review current approaches to these issues, including virtual machines
and workflow systems, and their limitations. I then examine how the
popular emerging technology Docker combines several areas from systems
research - such as operating system virtualization, cross-platform
portability, modular re-usable elements, versioning, and a `DevOps'
philosophy, to address these challenges. I illustrate this with several
examples of Docker use with a focus on the R statistical environment.
\end{abstract}

% A category with the (minimum) three required fields
%\category{H.4}{Information Systems Applications}{Miscellaneous}
%A category including the fourth, optional field follows...
%\category{D.2.8}{Software Engineering}{Metrics}[complexity measures, performance measures]

%\terms{Systems, Reproducible Research}
%

\section{Introduction}\label{introduction}

Reproducible research has received an increasing level of attention
throughout the scientific community {[}19, 22{]} and the public at large
{[}25{]}. All steps of the scientific process, from data collection and
processing, to analyses, visualizations and conclusions depend ever more
on computation and algorithms, \emph{computational reproducibility} has
received particular attention {[}18{]}. Though in principle this
algorithmic dependence should make such research easier to reproduce --
computer codes being both more portable and potentially more precise to
exchange and run than experimental methods -- in practice this has led
to an ever larger and more complex black box that stands between what
was actually done and what is described in the literature. Crucial
scientific processes such as replicating the results, extending the
approach or testing the conclusions in other contexts, or even merely
installing the software used by the original researchers can become
immensely time-consuming if not impossible.

\subsubsection{Systems research \&
reproducibility}\label{systems-research-reproducibility}

Systems research has long concerned itself with the issues of
computational reproducibility and the technologies that can facilitate
those objectives {[}6{]}. Docker is a new but already very popular open
source tool that combines many of these approaches in a user friendly
implementation, including: (1) performing Linux container (LXC) based
operating system (OS) level virtualization, (2) portable deployment of
containers across platforms, (3) component reuse, (4) sharing, (5)
archiving, and (6) versioning of container images. While Docker's market
success has largely focused on the needs of businesses in deploying web
applications and the potential for a lightweight alternative to full
virtualization, these features have potentially important implications
for systems research in the area of scientific reproducibility.

In this paper, I seek to address two audiences. First, that of the
domain scientist, those conducting research in ecology, bioinformatics,
economics, psychology and so many other disciplines in which computation
plays an ever-increasing role. I seek to help this audience become more
aware of the concerns and challenges in making these computations more
reproducible, extensible, and portable to other researchers, and
highlight Docker as platform to address these challenges. The second
audience is that of the computer systems researchers: readers familiar
with both the challenges to reproducibility (versioning, software
dependencies, etc.) and the technical components underlying Docker
(virtualization, LXC containers, jails, hashes, etc.), but who may be
unfamiliar with Docker software. I hope the domain scientist will be
motivated to try Docker as a tool to address reproducibility. Meanwhile,
I expect the systems researcher to see Docker technology as an area ripe
for further systems research (e.g.~to what extent can differences
between Linux kernels frustrate reproducibility in a Docker image?)
rather than as a tool for their own study (because Docker applies only
to software at the level above the Linux kernel, it will be of little
use in making studies of differences in kernel behavior, hardware
performance, execution times, and so forth reproducible.)

\subsubsection{A cultural problem}\label{a-cultural-problem}

It is worth observing from the outset that the primary barrier to
computational reproducibility in many domain sciences has nothing to do
with the technological approaches discussed here, but stems rather from
a reluctance to publish the code used in generating the results in the
first place {[}2{]}. Despite extensive evidence to the contrary
{[}14{]}, many researchers and journals continue to assume that summary
descriptions or pseudo-code provide a sufficient description of
computational methods used in data gathering, processing, simulation,
visualization, or analysis. Until such code is available in the first
place, one cannot even begin to encounter the problems that the
approaches discussed here set out to solve. As a result, few domain
researchers may be fully aware of the challenges involved in effectively
re-using published code.

A lack of requirements or incentives no doubt plays a crucial role in
discouraging sharing {[}2, 24{]}. Nevertheless, it is easy to
underestimate the significant barriers raised by a lack of familiar,
intuitive, and widely adopted tools for addressing the challenges of
computational reproducibility. Surveys and case studies find that a lack
of time, more than innate opposition to sharing, discourages researchers
from providing code {[}7, 23{]}.

\subsection{Four technical challenges}\label{four-technical-challenges}

@@

Studies focusing on code that has been made available with scientific
publications regularly find the same common issues that pose substantial
barriers to reproducing the original results or building on that code
{[}4, 8, 10, 16, 27{]}, which I attempt to summarize as follows.

\subsubsection{1. \textbf{``Dependency Hell''}}\label{dependency-hell}

A recent study by researchers at the University of Arizona found that
less than 50\% of software could even be successfully built or installed
{[}4{]}. Though sufficiently knowledgeable users may be able to overcome
the issues in at least some of these cases,\footnote{See discussion at
  \url{https://gist.github.com/pbailis/9629050} and
  \url{https://gist.github.com/samth/9641364}} similar results are seen
in an ongoing effort by other researchers to replicate that study
{[}27{]}, and has also been observed in other indepedent studies {[}8,
16{]}. Installing or building software necessary to run the code in
question assumes the ability to recreate the computational environment
of the original researchers.

Differences in numerical evaluation, such as arise in floating point
arithmetic or even ambiguities in standardized programming languages
(``order-of-evaluation'' problems) can be responsible for differing
results between or even within the same computational platform {[}14{]}.
Such issues make it difficult to restrict the true dependencies of the
code to higher level environments such as that of a given scripting
language, independent of the underlying OS or even hardware itself.

\subsubsection{2. \textbf{Imprecise
documentation}}\label{imprecise-documentation}

Documentation on how to install and run code associated with published
research is another frequent barrier to replication. A study by Lapp
{[}16{]} found this impairs a researcher's ability to install and build
the software necessary, as even small holes in the documentation were
found to be major barriers, particularly for ``novices'' {[}8{]} --
where novices may be experts in nearby languages but unfamiliar with the
package managers and other tools of the language involved. This same
problem is discussed in {[}3{]}. Imprecise documentation goes well
beyond issues of the software environment itself: incomplete
documentation of parameters involved meant as few as 30\% of analyses
(\(n=34\)) using the popular software STRUCTURE could be reproduced in
the study of {[}10{]}.

\subsubsection{3. \textbf{Code rot}}\label{code-rot}

Software dependencies are not static elements, but receive regular
updates that may fix bugs, add new features or deprecate old features
(or even entire dependencies themselves). Any of these changes can
potentially change the results generated by the code. As some of these
changes may indeed resolve valid bugs or earlier problems with
underlying code, it will often be insufficient to demonstrate that
results can be reproduced when using the original versions, a problem
sometimes known as ``code rot.'' Researchers will want to know if the
results are robust to the changes. The case studies in {[}16{]} provide
examples of these problems.

\subsubsection{4. \textbf{Barriers to adoption and reuse in existing
solutions}}\label{barriers-to-adoption-and-reuse-in-existing-solutions}

Technological solutions such as workflow software, virtual machines,
continuous integration services, and best practices from software
development would address many of the issues frequently frustrating
reproducibility. However, researchers face significant barriers to entry
in learning these tools and approaches which are not part of their
typical curriculum, or lack incentives commensurate with the effort
required {[}7, 15{]}.

Though a wide variety of approaches exists to work around these
challenges, few operate on a low enough level to provide a general
solution. Clark \emph{et al.} {[}3{]} provide an excellent description
of this situation:

\begin{quote}
In scientific computing the environment was commonly managed via
Makefiles \& Unix-y hacks, or alternatively with monolithic software
like Matlab. More recently, centralized package management has provided
curated tools that work well together. But as more and more essential
functionality is built out across a variety of systems and languages,
the value -- and also the difficulty -- of coordinating multiple tools
continues to increase. Whether we are producing research results or web
services, it is becoming increasingly essential to set up new languages,
libraries, databases, and more.
\end{quote}

@@

\section{Current approaches}\label{current-approaches}

Two dominant paradigms have emerged to address these issues so far:
workflow software {[}1, 13{]} and virtual machines {[}5, 12{]}. Workflow
software provides very elegant technical solutions to the challenges of
communication between diverse software tools, capturing provenance in
graphically driven interfaces, and handling issues from versioning
dependencies to data access. Workflow solutions are often built by
well-funded collaborations between domain scientists and computer
scientists, and can be very successful in the communities within which
they receive substantial adoption. Nonetheless, most workflow systems
struggle with relatively low total adoption overall {[}5, 9{]}.

Dudley \& Butte {[}5{]} give several reasons that such comprehensive
workflow systems have not been more successful:

\begin{quote}
\begin{enumerate}
\def\labelenumi{(\roman{enumi})}
\itemsep1pt\parskip0pt\parsep0pt
\item
  efforts are not rewarded by the current academic research and funding
  environment; (ii) commercial software vendors tend to protect their
  markets through proprietary formats and interfaces; (iii)
  investigators naturally tend to want to `own' and control their
  research tools; (iv) even the most generalized software will not be
  able to meet the needs of every researcher in a field; and finally (v)
  the need to derive and publish results as quickly as possible
  precludes the often slower standards-based development path.
\end{enumerate}
\end{quote}

In short, workflow software expects a new approach to computational
research. In contrast, virtual machines (VMs) offer a more direct
approach. Since the computer Operating System (OS) already provides the
software layer responsible for coordinating all the different elements
running on the computer, the VM approach captures the OS and everything
running on it whole-cloth. To make this practical, Dudley \& Butte
{[}5{]} and Howe {[}12{]} both propose using virtual machine images that
will run on the cloud, such as Amazon's EC2 system, which is already
based upon this kind of virtualization.

Critics of the use of VMs to support reproducibility highlight that the
approach is too much of a black box and thus ill suited for
reproducibility {[}28{]}. While the approach sidesteps the need to
either install or even document the dependencies, this also makes it
more difficult for other researchers to understand, evaluate, or alter
those dependencies. Moreover, other research cannot easily build on the
virtual machine in a consistent and scalable way. If each study provided
its own virtual machine, any pipeline combining the tools of multiple
studies would quickly become impractical or impossible to implement.

\subsection{A ``DevOps'' approach}\label{a-devops-approach}

The problems highlighted here are not unique to \emph{academic}
software, but impact software development in general. While the academic
research literature has frequently focused on the development of
workflow software dedicated to particular domains, or otherwise to the
use of virtual machines, the software development community has recently
emphasized a philosophy (rather than a particular tool), known as
\emph{Development} and Systems \emph{Operation}, or more frequently just
``DevOps.'' The approach is characterized by scripting, rather than
documenting, a description of the necessary dependencies for software to
run, usually from the Operating System (OS) on up. Clark \emph{et al.}
{[}3{]} describe the DevOps approach along with both its relevance to
reproducible research and examples of its use in the academic research
context. They identify the difficulties I have discussed so far in terms
of effective documentation:

\begin{quote}
Documentation for complex software environments is stuck between two
opposing demands. To make things easier on novice users, documentation
must explain details relevant to factors like different operating
systems. Alternatively, to save time writing and updating documentation,
developers like to abstract over such details.
\end{quote}

The authors contrast this to the DevOps approach, where dependency
documentation is \emph{scripted}:

\begin{quote}
A DevOps approach to ``documenting'' an application might consist of
providing brief descriptions of various install paths, along with
scripts or ``recipes'' that automate setup.
\end{quote}

This elegantly addresses both the demand for simplicity of use (one
executes a script instead of manually managing the environmental setup)
and comprehensiveness of implementation. Clark \emph{et al.} {[}3{]} are
careful to note that this is not so much a technological shift as a
philosophical one:

\begin{quote}
The primary shift that's required is not one of new tooling, as most
developers already have the basic tooling they need. Rather, the needed
shift is one of philosophy.
\end{quote}

Nevertheless, a growing suite of tools designed explicitly for this
purpose have rapidly replaced the use of general purpose tools (such as
Makefiles, bash scripts) to become synonymous with the DevOps
philosophy. Clark \emph{et al.} {[}3{]} reviews many of these DevOps
tools, their different roles, and their application in reproducible
research.

I focus the remainder of this paper on one of the most recent and
rapidly growing among these, called Docker, and the role it can play in
reproducible research. Docker offers several promising features for
reproducibility that go beyond the tools highlighted in {[}3{]}.
Nevertheless, my goal in focusing on this technology is not to promote a
particular solution, but to anchor the discussion of technical solutions
to reproducibility challenges in concrete examples.

\section{Docker}\label{docker}

Docker is an open source project that builds on many long-familiar
technologies from operating systems research: LXC containers,
virtualization of the OS, and a hash-based or git-like versioning and
differencing system, among others (see
\href{https://docs.docker.com/faq/\#what-does-docker-add-to-just-plain-lxc}{docs.docker.com/faq}
for an excellent overview of what Docker adds to plain LXC). The
official documentation \href{https://docs.docker.com}{docs.docker.com}
already provides a thorough introduction in how to use Docker software;
here my focus is on describing the implications this has for
reproducible research. Readers may also find it helpful to see more
detailed examples of using Docker to capture, share, and interact with a
specific computational environments. Some such examples, along with more
detailed documentation on use can be found at
\href{https://github.com/rocker-org/rocker}{github.com/rocker-org}.
